On Verification Patterns for Low-Level Systems via Modal Abstractions

### 2025-09-11 paper.tex
+

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle
\section{Introduction}
Although they differ in the functionality they offer, low-level systems exhibit certain patterns of design and utilization of computing resources.
Systems software, in general, interfaces with an underlying computing substrate such that any software system at any higher level in the software stack can (at least indirectly) utilize the resources of the machine.
The last layer of software before the hardware is naturally critical to the correctness of an overall system,
as essentially all software built on top of it assumes its correctness.
%Assuming verified hardware, this makes the correctness of the system software critical to the rest of the software stack. 
%Because they interface with the hardware, abstractions and their implementations are intricate and error-prone. 
And because hardware is complex and highly diverse, the implementation of those lowest layers
of the software stack is typically intricate and naturally error-prone, despite how critical its correctness is.
Typically systems software has, as a primary focus, the task of \emph{abstracting} from hardware details to simplify 
the construction of higher layers of the stack.

@@

\section{Background on Modal Logics for Programs, and Low-Level Systems Software}

This section briefly recalls related concepts from modal logic of particular relevance
to our taxonomy, which should be familiar but benefit from fixing notation (Section \ref{sec:modalbackground}),
and enough general background on the systems software concepts organized by our taxonomy
for the connections to modalities to be clear (Section \ref{sec:background}).

\subsection{Hybrid Logic, Dynamic Logic, and Nominals in Program Logics}
\label{sec:modalbackground}
\textit{Dynamic Logics as Program Logics}
We believe that for the kinds of informal reasoning and the sorts of data structures discussed in the previous section, ideas from modal logic are a promising approach to formalization and verification.
Broadly speaking, modal logics incorporate \emph{modal operators}, which take as arguments a proposition expected to be true in another time~\cite{pnueli1977temporal}, place~\cite{gordon2019modal,goranko1996hierarchies,areces2001hybrid,gargov1993modal,murphy2004symmetric}, circumstance or point-of-view~\cite{hintikka1962knowledge,halpern1985guide}, and result in a proposition true in the \emph{current} time, place, or circumstance at which the truth of the use of the modal operator is being evaluated. Classic examples include modal necessity $\square P$ describing that $P$ is \emph{necessarily} true, $\mathsf{G}~P$ meaning $P$ is true \emph{globally} (i.e., forever from this time onwards in temporal logics), or $K_i(P)$ describing that a particular participant $i$ \emph{knows} that $P$ is true. The latter is an example of \emph{multimodal} logic, where there is an family of modalities (modal operators) parameterized by some dimension of interest (there, participants).

\emph{Dynamic logic}~\cite{pratt1976semantical} is a (multi)modal logic of weakest preconditions~\cite{dijkstra-75} which works with modalities of the form $[p](P)$, which states that \emph{in the current program state}, \emph{if} program $p$ is run then afterwards $P$ will hold (modulo non-termination).
This same idea is used to encode Hoare triples in the Iris program logic~\cite{jung2018iris}, using the same encoding as in Pratt's original presentation, where a Hoare triple $\{P\}C\{Q\}$ is encoded as $P\rightarrow[C](Q)$.
Iris is now one of the most successful foundational program logics, and its rich support for
extending the base logic with additional modalities~\cite{krebbers2017essence} means that
modal logics, and in particular dynamic logic, now play an outsized role in program verification ---
many challenging proofs of program correctness use a substructural dynamic logic containing
many additional modalities which interact with the weakest precondition modality.

@@

\textit{Hybrid Logic} Hybrid logics~\cite{blackburn1995hybrid,goranko1996hierarchies,gargov1993modal,areces2001hybrid,brauner2010hybrid} are a class of modal logics that add two primary new concepts to a logic. Nominals $\ell \in \textsf{Loc}$ uniquely identify points in a model (e.g., a particular state), so there is exactly one point in the model where a nominal $\loc$ is true. Satisfaction operators are modal operators indexed by nominals, which enable claims about the truth of another proposition at some arbitrary point in the model identified by a nominal. Traditionally $@_l\varphi$ asserts that $\varphi$ is true in the state identified by $l$. Conceptually, given our context, we might like $l$ to indicate truth in a particular data structure $l$.  However, most prior work on hybrid logics works in classical logics rather than a substructural setting like separation logic. 
Newer work ~\cite{tejthesis,perennialgit,tejperennial19,larsnextgen25,fsl,fsl++,derekrustbelt20,kuru2025modalabstractionsvirtualizingmemory,amalreal2024}  surveyed here adapts this idea to deal with different structures specified in
separation logic.
%We adapt these ideas to assertions of the form $@[l]\varphi$ asserts that $\varphi$ is true in a data structure with \emph{logical} state $\sigma$ \emph{at location} $l$.
% In practice we must design variants of this for each kind of structure we wish to apply this to, e.g. tree nodes, free list, or snapshots.
%This section provides details of and models for each modal abstraction, focusing on how each data structure can act as a model for a data-structure-specific logic.
%We continue to motivate the development with aspects of the write amplification process described in the previous section; Section \ref{sec:evaluation} provides further examples using the modal abstractions in proofs.
% \paragraph{Temporal Operators}
% The best known class of modal logics in the systems community is undoubtedly temporal logics (whether \textsc{LTL}~\cite{pnueli1977temporal}, \textsc{CTL}~\cite{emerson1982using}, \textsc{TLA}~\cite{lamport1994temporal}, or others),
% due in part to Lamport's influence~\cite{lamport2002specifying} leading to its not-infrequent use in specifying
% distributed algorithms \cite{ongaro2014search,ailijiang2019wpaxos}.

\textit{Nominalization}
Some classes of assertions benefit specifically from \emph{naming} the explicit conditions where they are true (as opposed to simply requiring them to be true \emph{somewhere} or \emph{everywhere} as in the most classic modal operators).
% This naming generally resembles
% the \emph{satisfaction} operator of 
% \emph{hybrid logic}~\cite{blackburn1995hybrid,areces2001hybrid}: $@_\iota P$ which evaluates the truth of $P$ at the named (Kripke model) state $\iota$. 
Because these named conditions strongly resemble the nominals of hybrid logic,
we refer to the general idea of naming circumstances explicitly as \emph{nominalization},
even though not all of the examples we discuss are actually hybrid logics.

An example use of state naming explicitly in the assertion appears in program logics 
with usage protocols (e.g. state transition systems resembling typestate~\cite{strom1986typestate,garcia2014foundations}, which ensure
the operations applied to some data occur in a specified order, and that invariants for various conceptual states of the data
--- such as a file being open or closed --- are respected). 
Protocol assertions are \emph{annotated with the name of the last (abstract) state at which the protocol is ensured}.
A classic application of this idea to systems is 
Halpern et al.'s work adapting modal logics of
knowledge to deal with distributed systems~\cite{halpern1990knowledge,halpern1989modelling,halpern2017reasoning}.
In most of that line of work, $\mathcal{K}_a(P)$ indicates that the node $a$ in the system \emph{knows} or \emph{possesses knowledge of} $P$
(for example, a Raft node may ``know'' a lower bound on the commit index).
Alternatively, a modality $@_i(P)$ may represent that $P$ is true of/at the specific node $i$~\cite{gordon2019modal} (e.g., that node $i$ has stored a certain
piece of data to reliable storage).
These permit capturing specific concepts relevant to the correctness (and reasoning about correctness) of a certain class of systems,
involving facts about specific named entities in the system.

In each of these cases, the fact that these facts are described using modalities with the core modal property $P \Rightarrow Q$, then $M(P) \Rightarrow M(Q)$ means, for example, that if a
process $p$ knows that the commit index is greater than 5 (e.g., $\mathcal{K}_p(\mathsf{commitIndex} > 5)$)
no extra work is required to conclude that the node knows it is greater than 3 (i.e., that $\mathcal{K}_p(\mathsf{commitIndex} > 3)$),
because this follows from standard properties of modal operators as described above.
In contrast, if verification instead used a custom assertion $\mathsf{minCommitIndex}(5)$ to represent the former knowledge,
one would need to separately provide custom reasoning to conclude $\mathsf{minCommitIndex}(3)$.
As our survey shows, the value of the modal view (which grants both intuitive specifications and natural
reasoning principles) is becoming apparent in systems software verification.

@@

\textit{Access by Translation}
Accessing virtualized resources via translation is a common way to virtualize notions of location (e.g.,
virtual memory addresses, inodes or object IDs instead of disk addresses~\cite{bonwick2003zettabyte,hitz1994file,Rosenblum1992LFS,rodeh2013btrfs}).
%Accessing the hardware resources in the presence of sharing means having more resource references (references to the virtualized resources) to access the \emph{mapped} resources that are physical and potentially fewer. 
B-trees, page tables, and related structures both behave like maps, when the corresponding physical resources exist as such
just in a different location.
Control over the lookup process (e.g, handling the case of a missing translation entry) allows for additional flexibility,
such as filling holes in sparse files, or demand paging (both from disk or lazily populating anonymous initially-zeroed mappings).
Although the realization of these maps may differ from system to system based on the context (and sometimes hardware details),
they are semantically --- logically --- partial maps, worth treating as such in verification.

\subsubsection{Virtualization of Memory Locations}
\label{sec:backgroundonmachinemodel}
A typical general purpose computer virtualizes  memory resources in RAM.
A program asking for a memory unit from an operating system kernel is served with a memory address that is \emph{virtualized}
such that it may be \emph{mapped to} a physical RAM location or not. 
As our informal definition suggests, the common technique for \emph{virtualization} of memory locations in OS 
kernels relies \emph{translating} (\emph{mapping}) virtual addresses into physical ones. 
The conceptual address-translation map is implemented with processor's page-table trees. A virtual address indexing into the different levels of the page-table tree 
are blocks of memory which serve as nodes of the tree along a particular lookup path. 
Fabrication of virtual addresses out of limited physical addresses is provided with entries 
in the page-table tree where each points to a physical address aligned to 4KB (4096 byte) boundaries. 
An address translation requires traversal of a series of tables starting from top level, and it is  conventionally referred as a \emph{page-table walk}. 
The traversal ends in the level 1 page table with the final lookup to the actual page of physical memory holding the requested data, and the low-order 12 bits being used to index into this page.
%\begin{figure}
%    \includegraphics[width=\columnwidth]{pagetables2.png}
%    \caption{x86-64 page table lookups.}
%    \label{fig:pagetables}
%\end{figure}

@@

% \paragraph{Virtual Memory Managers(VMM) with Multiple Address-Spaces}
% \label{sec:backgroundonvmm}
% \todo{change the text}
% An intuition for the abstraction that \emph{contains} a page-table tree is called an \emph{address-space} of a process whose virtual addresses to physical addresses. 
A typical OS manages multiple address spaces at once.
Each program is associated with a unique page table tree, which is stored in a specific hardware register (e.g., \texttt{cr3}). 
Using different mappings, which map only disjoint portions of physical memory in the map's codomain, 
is how the OS ensures memory isolation between processes.

\subsubsection{Virtualized Resources in Filesystems}
Another important computer resource computer resource that is \emph{virtualized} for access is disk blocks. 
Filesystems are software components that abstract disk access on behalf of normal programs.
Similarly to how VMMs use page-table trees to fabricate more memory addresses than actually exist, filesystems use indexing maps to fabricate more virtual disk-page addresses (which are conventionally called disk-pages) to be distinguished from physical disk-block addresses.
When a regular program accesses a file, it requests a specific logical offset from the start of that file's contents.
As with virtual memory, the filesystem may map adjacent logical blocks of a single file to physically disparate parts of the disk (or even
to different disks), though this translation mapping is implemented entirely in software (as opposed to the hardware cooperation that exists
for memory virtualization).

% \paragraph{Overview of Policies in Filesystems}
Filesystems not only handle the address-translation from a disk-page to a disk-block but also need to handle different modes of computing, representing where to find disk resources: a disk block may live on a disk in stable storage,
or the most up-to-date version of a block may live only in memory (for the filesystem to write to disk at a later time, batched for efficiency).
To handle the consistency of the disk resources in the existence of different modes of execution, filesystems employ different policies when indexing the disk resource.
Which data-structure is going to be used to implement address-translation from disk-pages to disk-blocks? 
What is going to be the \emph{update-policy} to the indexing map implementing address-translation which defines the \emph{consistency-policy} for the disk resources? 
Different filesystem implementations give different answers to these questions.

@@

\subsubsection{Reference-Counting Memory Reclamation}
In a typical configuration, allocating, deallocating and referencing memory locations for client code 
leaves the memory layout \emph{ideally} in a form of DAG (directed acyclic graph) with a unique root, which resembles the first stack pointer referencing the allocated memory. DAG abstraction instead of recursive descent tree allows non-duplication on the reference counts for the identical memory objects.
% To ease picturing the memory layout after a set of requests from the client code (for allocation, referencing etc.), we can think of the unique roots of DAGs as 
% \emph{super} block addresses -- this time multiple ones, not only two -- in Figure \ref{fig:cowindex}, 
% from which a path is created for a stack pointer reference to the allocated memory. 
%\todo[inline,color=cyan]{
%Why is this a DAG? Normally programs can easily create cycles, so the DAG bit requires a little more explanation.
%Do we mean that it is ``often'' the case that it's a DAG?}

@@

\textit{Shallow Modalities in the Iris Separation Logic}
\label{sec:modal_in_iris}
Classically modal operators are given direct semantics in terms of an underlying model, 
which for \textsf{Iris}~\cite{jung2018iris} would be a collection of \textsc{CMRA}s (complete metric resource algebras) ---
an algebraic cousin of a BI-algebra~\cite{ohearn1999bunched}.
This is the case for \textsf{Iris}'s one built-in modality for step-indexing~\cite{nakano-00}.
However, in general, many modalities can be given proof-theoretic semantics directly within a logic. 
This is the case for \textsf{Iris}'s weakest precondition modality (which is defined within the \textsf{Iris} 
logic itself), as well as for a wide range of other modalities in the 
literature~\cite{restall2002introduction}. These modalities are typically multi-modal, 
with a form of roughly $M~\overline{ctxt}~\phi\triangleq\ldots\rightarrow \phi~\overline{ctxt}$. 
In a substructural setting~\cite{restall1993modalities,krebbers2017essence} the implication $\rightarrow$ 
is typically replaced with substructural implication $\sepimp$.

This definition also reveals a subtlety about the denotation of assertions $\phi$ --- they are effectively \emph{predicates} of some kind, functions from some possible world into some logical algebra such as a BI-algebra~\cite{ohearn1999bunched}. 
Taking this interpretation literally gives rise to a convenient way to embed new modalities in \textsf{Iris} 
in a way well-suited to many use cases. 
If $B$ is a type with a BI-algebra structure, then any space of predicate functions (e.g., $\mathsf{val}\rightarrow B$ as a simplification) also carries a BI-algebra structure, lifted from $B$, often called the 
\emph{pointwise lifting} of the algebra. A number of 
projects for systems software verification ~\cite{tejthesis,perennialgit,tejperennial19,larsnextgen25,fsl,fsl++,derekrustbelt20,kuru2025modalabstractionsvirtualizingmemory,amalreal2024} now take advantage of this,
explicitly or implicitly lifting the BI-algebra structure from \textsf{Iris} itself to \textsf{Iris} predicates 
used as assertions within modalities. 
Connecting this to \textsf{Iris}'s typeclasses for interactive proofs~\cite{krebbers2018mosel,krebbers-17} then permits working within the embedded logic as in \textsf{Iris}.

@@

\subsection{Decomposing a System into its Constituents \emph{Contingently}}
\label{sec:definitions}
\begin{table*}[t]\scriptsize
\centering
\caption{Modal Decomposition of Program-Logics.}
\label{table:decomposition}
 %$\{\Diamond (P) \}$ write  $\{ P\}_{n}$
%$\{\overset{ICut^{n}}{\hookrightarrow} (P) \}$ return $\{ P\}_{n} $
\begin{tabular}{@{}lcccp{4cm}@{}}
  \toprule
{ Modality}& { Context} & { Elements}  &  { Nominalization} & { Context Steps} \\ \midrule
 Post-Crash~\cite{tejthesis,perennialgit,tejperennial19} &  $\Diamond \; P $  &  $  \ell \mapsto_{n}^{\overline{\gamma}} v $ &  Strong &   Crash Recovery    \\ 
%GC-Modality &      & \multicolumn{2}{c}{\multirow{2}{*}{Because why not}} & 8                                   & & ~\cite{}\\
 NextGen~\cite{larsnextgen25} &   $\overset{t}{\hookrightarrow} \; P$  &  Own (t(a))  &  Strong &  Determined Based on the Model$^{*}$   \\
 StackRegion$^{*}$  ~\cite{larsnextgen25} &  $\overset{ICut^{n}}{\hookrightarrow} \; P$      &  $\fbox{n} \; \ell \mapsto v$                                 &  Strong &  Alloc and Return to/from stack\\
 Memory-Fence~\cite{fsl,fsl++,derekrustbelt20} &  $\triangle_{\pi}$ and $\triangledown_{\pi}$      &  $\ell \mapsto v$                                 &  Weak &  Fence Acquire and Release  \\
 Address-Space~\cite{kuru2025modalabstractionsvirtualizingmemory} &  [r]P     &   $\ell \mapsto v$ &  Weak                         &  Address-Space Switch  \\ 
 Ref-Count~\cite{amalreal2024}& @$_{\ell}$ P    &  $\ell_1 \mapsto v$ &  Weak  &  Allocating, Dropping and Sharing a Reference \\ \bottomrule
\end{tabular}

{\scriptsize *The StackRegion Modality is an instance of NextGen (called the Independence Modality in \cite{larsnextgen25})}.
%and the $\Diamond$ Figure for Post-Crash Modality uses notationis taken from NextGen Modality paper ~\cite{larsnextgen25}.
\end{table*}

@@

\begin{definition}[Resource Elements]
%Inherited from separation logic, we identify the resource element as the indivisible logical element on which contingency defined by the modality holds. 
The simplest atomic facts we want to work with in a particular setting, specific to that setting.
\end{definition}
By definition, the resource elements are specific to some limited domain or setting.
For example, knowing that a certain address points to a 32-bit signed integer representing 3 is knowledge restricted to a certain address space.
In general,  we call these domains that any resource element is tied to \emph{resource contexts}:
\begin{definition}[Resource Context]
A resource context is an abstraction, context, or container of resource elements of the same type, e.g., an address space of a process.
\end{definition}
We discuss a range of examples for each of these in turn. Table \ref{table:decomposition} gives additional examples of systems and their corresponding resource elements and contexts where these elements reside, though none of the work in that table analyzes itself according to the structure we are giving.

>>>>

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle
\section{Introduction}
Although they differ in the functionality they offer, low-level systems exhibit certain patterns of design and utilization of computing resources.
Systems software, in general, interfaces with an underlying computing substrate such that any software system at any higher level in the software stack can (at least indirectly) utilize the resources of the machine.

The last layer of software before the hardware is naturally critical to the correctness of an overall system,
as essentially all software built on top of it assumes its correctness.
%Assuming verified hardware, this makes the correctness of the system software critical to the rest of the software stack. 
%Because they interface with the hardware, abstractions and their implementations are intricate and error-prone. 
And because hardware is complex and highly diverse, the implementation of those lowest layers
of the software stack is typically intricate and naturally error-prone, despite how critical its correctness is.
Typically systems software has, as a primary focus, the task of \emph{abstracting} from hardware details to simplify 
the construction of higher layers of the stack.

@@

\subsection{Hybrid Logic, Dynamic Logic, and Nominals in Program Logics}
\label{sec:modalbackground}

\paragraph{Dynamic Logics as Program Logics}
We believe that for the kinds of informal reasoning and the sorts of data structures discussed in the previous section, ideas from modal logic are a promising approach to formalization and verification.
Broadly speaking modal logics incorporate \emph{modal operators}, which take as arguments a proposition expected to be true in another time~\cite{pnueli1977temporal}, place~\cite{gordon2019modal,goranko1996hierarchies,areces2001hybrid,gargov1993modal,murphy2004symmetric}, circumstance or point-of-view~\cite{hintikka1962knowledge,halpern1985guide}, and result in a proposition true in the \emph{current} time, place, or circumstance at which the truth of the use of the modal operator is being evaluated. Classic examples include modal necessity $\square P$ describing that $P$ is \emph{necessarily} true, $\mathsf{G}~P$ meaning $P$ is true \emph{globally} (i.e., forever from this time onwards in temporal logics), or $K_i(P)$ describing that a particular participant $i$ \emph{knows} that $P$ is true. The latter is an example of \emph{multimodal} logic, where there is an family of modalities (modal operators) parameterized by some dimension of interest (there, participants).

\emph{Dynamic logic}~\cite{pratt1976semantical}, a (multi)modal logic of weakest preconditions~\cite{dijkstra-75} which works with modalities of the form $[p](P)$, which states that \emph{in the current program state}, \emph{if} program $p$ is run then afterwards $P$ will hold (modulo non-termination).
This same idea is used to encode Hoare triples in the Iris program logic~\cite{jung2018iris}, using the same encoding as in Pratt's original presentation, where a Hoare triple $\{P\}C\{Q\}$ is encoded as $P\rightarrow[C](Q)$.
Iris is now one of the most successful foundational program logics, and its rich support for
extending the base logic with additional modalities~\cite{krebbers2017essence} means that
modal logics, and in particular dynamic logic, now play an outsized role in program verification ---
many challenging proofs of program correctness use a substructural dynamic logic containing
many additional modalities which interact with the weakest precondition modality.

@@

\paragraph{Hybrid Logic} Hybrid logics~\cite{blackburn1995hybrid,goranko1996hierarchies,gargov1993modal,areces2001hybrid,brauner2010hybrid} are a class of modal logics that add two primary new concepts to a logic. Nominals $\ell \in \textsf{Loc}$ uniquely identify points in a model (e.g., a particular state), so there is exactly one point in the model where a nominal $\loc$ is true. Satisfaction operators are modal operators indexed by nominals, which enable claims about the truth of another proposition at some arbitrary point in the model identified by a nominal. Traditionally $@_{\ell}\varphi$ asserts that $\varphi$ is true in the state identified by $\ell$. Conceptually, given our context, we might like $l$ to indicate truth in a particular data structure $\ell$.  However, most prior work on hybrid logics works in classical logics rather than a substructural setting like separation logic. 
Newer work surveyed here adapts this idea to deal with different structures specified in
separation logic.
%We adapt these ideas to assertions of the form $@[l]\varphi$ asserts that $\varphi$ is true in a data structure with \emph{logical} state $\sigma$ \emph{at location} $l$.
% In practice we must design variants of this for each kind of structure we wish to apply this to, e.g. tree nodes, free list, or snapshots.
%This section provides details of and models for each modal abstraction, focusing on how each data structure can act as a model for a data-structure-specific logic.
%We continue to motivate the development with aspects of the write amplification process described in the previous section; Section \ref{sec:evaluation} provides further examples using the modal abstractions in proofs.

% \paragraph{Temporal Operators}
% The best known class of modal logics in the systems community is undoubtedly temporal logics (whether \textsc{LTL}~\cite{pnueli1977temporal}, \textsc{CTL}~\cite{emerson1982using}, \textsc{TLA}~\cite{lamport1994temporal}, or others),
% due in part to Lamport's influence~\cite{lamport2002specifying} leading to its not-infrequent use in specifying
% distributed algorithms \cite{ongaro2014search,ailijiang2019wpaxos}.

\paragraph{Nominalization}
Some classes of assertions benefit specifically from \emph{naming} the explicit conditions where they are true (as opposed to simply requiring them to be true \emph{somewhere} or \emph{everywhere} as in the most classic modal operators).
% This naming generally resembles
% the \emph{satisfaction} operator of 
% \emph{hybrid logic}~\cite{blackburn1995hybrid,areces2001hybrid}: $@_\iota P$ which evaluates the truth of $P$ at the named (Kripke model) state $\iota$. 
Because these named conditions strongly resemble the nominals of hybrid logic,
we refer to the general idea of naming circumstances explicitly as \emph{nominalization},
even though not all of the examples we discuss are actually hybrid logics.

An example use of state naming explicitly in the assertion appears in program logics 
with usage protocols (e.g. state transition systems resembling typestate~\cite{strom1986typestate,garcia2014foundations}, which ensure
the operations applied to some data occur in a specified order, and that invariants for various conceptual states of the data
--- such as a file being open or closed --- are respected). 
Protocol assertions are \emph{annotated with the name of the last (abstract) state at which the protocol is ensured}.
A classic application of this idea to systems is 
Halpern et al.'s work adapting modal logics of
knowledge to deal with distributed systems~\cite{halpern1990knowledge,halpern1989modelling,halpern2017reasoning}.
In most of that line of work, $\mathcal{K}_a(P)$ indicates that the node $a$ in the system \emph{knows} or \emph{possesses knowledge of} $P$
(for example, a Raft node may ``know'' a lower bound on the commit index).
Alternatively, a modality $@_i(P)$ may represent that $P$ is true of/at the specific node $i$~\cite{gordon2019modal} (e.g., that node $i$ has stored a certain
piece of data to reliable storage).
These permit capturing specific concepts relevant to the correctness (and reasoning about correctness) of a certain class of systems,
involving facts about specific named entities in the system.
\\

In each of these cases, the fact that these facts are described using modalities with the core modal property $P \Rightarrow Q$, then $M(P) \Rightarrow M(Q)$ means, for example, that if a
process $p$ knows that the commit index is greater than 5 (e.g., $\mathcal{K}_p(\mathsf{commitIndex} > 5)$)
no extra work is required to conclude that the node knows it is greater than 3 (i.e., that $\mathcal{K}_p(\mathsf{commitIndex} > 3)$),
because this follows from standard properties of modal operators as described above.
In contrast, if verification instead used a custom assertion \[ \mathsf{minCommitIndex}(5)\] to represent the former knowledge,
one would need to separately provide custom reasoning to conclude \[ \mathsf{minCommitIndex}(3)\].
As our survey shows, the value of the modal view (which grants both intuitive specifications and natural
reasoning principles) is becoming apparent in systems software verification.

@@

\paragraph{Access by Translation}
Accessing virtualized resources via translation is a common way to virtualize notions of location (e.g.,
virtual memory addresses, inodes or object IDs instead of disk addresses~\cite{bonwick2003zettabyte,hitz1994file,Rosenblum1992LFS,rodeh2013btrfs}).
%Accessing the hardware resources in the presence of sharing means having more resource references (references to the virtualized resources) to access the \emph{mapped} resources that are physical and potentially fewer. 
B-trees, page tables, and related structures both behave like maps, when the corresponding physical resources exist as such
just in a different location.
Control over the lookup process (e.g, handling the case of a missing translation entry) allows for additional flexibility,
such as filling holes in sparse files, or demand paging (both from disk or lazily populating anonymous initially-zeroed mappings).
Although the realization of these maps may differ from system to system based on the context (and sometimes hardware details),
they are semantically --- logically --- partial maps, worth treating as such in verification.

\subsubsection{Virtualization of Memory Locations}
\label{sec:backgroundonmachinemodel}
A typical general purpose computer virtualizes  memory resources in RAM.
A program asking for a memory unit from an operating system kernel is served with a memory address that is \emph{virtualized}
such that it may be \emph{mapped to} a physical RAM location or not. 
As our informal definition suggests, the common technique for \emph{virtualization} of memory locations in OS 
kernels relies \emph{translating} (\emph{mapping}) virtual addresses into physical ones. 
The conceptual address-translation map is implemented with processor's page-table trees as shown in 
Figure \ref{fig:pagetables}. In this Figure, we see a typical 4 levels of page tables, a virtual address on the 
top indexing into the different levels of the page-table tree (the larger rectangles along the bottom
are blocks of memory which serve as nodes of the tree along a particular lookup path). 
Fabrication of virtual addresses out of limited physical addresses is provided with entries 
in the page-table tree where each points to a physical address aligned to 4KB (4096 byte) boundaries. 
An address translation requires traversal of a series of tables starting from level 4, and it is 
conventionally referred as a \emph{page-table walk}. 
The traversal ends in the level 1 page table with the final lookup to the actual page of physical memory holding the requested data, and the low-order 12 bits being used to index into this page.
\begin{figure}
    \includegraphics[width=\columnwidth]{pagetables2.png}
    \caption{x86-64 page table lookups.}
    \label{fig:pagetables}
\end{figure}

@@

% \paragraph{Virtual Memory Managers(VMM) with Multiple Address-Spaces}
% \label{sec:backgroundonvmm}
% \todo{change the text}
% An intuition for the abstraction that \emph{contains} a page-table tree is called an \emph{address-space} of a process whose virtual addresses to physical addresses. 
A typical OS manages multiple address spaces at once.
Each program is associated with a unique page table tree, which is stored in a specific hardware register (\texttt{cr3} in Figure \ref{fig:pagetables}). 
Using different mappings, which map only disjoint portions of physical memory in the map's codomain, 
is how the OS ensures memory isolation between processes.

\subsubsection{Virtualized Resources in Filesystems}
Another important computer resource computer resource that is \emph{virtualized} for access is disk blocks. 
Filesystems are software components that abstract disk access on behalf of normal programs
Similar to how VMMs using page-table trees to fabricate more memory addresses than what actually exists, 
filesystems uses indexing maps for fabricating more \emph{virtual disk-page} addresses which are conventionally called \emph{disk-pages} 
to be distinguished from physical \emph{disk-block} addresses.
When a regular program accesses a file, it requests a specific logical offset from the start of that file's contents.
As with virtual memory, the filesystem may map adjacent logical blocks of a single file to physically disparate parts of the disk (or even
to different disks), though this translation mapping is implemented entirely in software (as opposed to the hardware cooperation that exists
for memory virtualization).

% \paragraph{Overview of Policies in Filesystems}
Filesystems not only handle the address-translation from a disk-page to a disk-block but also need to handle different
\emph{modes} of computing where to find disk resources: a disk block may live on a disk in stable storage,
or the most up-to-date version of a block may live only in memory (for the filesystem to write to disk at a later time, batched for efficiency).
To handle the consistency of the disk resources in the existence of different modes of execution, filesystems employ different policies when indexing the disk resource.
Which data-structure is going to be used to implement address-translation from disk-pages to disk-blocks? 
What is going to be the \emph{update-policy} to the indexing map implementing address-translation which defines the \emph{consistency-policy} for the disk resources? 
Different filesystem implementations give different answers to these questions.

@@

\subsubsection{Reference-Counting Memory Reclamation}
In a typical configuration, allocating, deallocating and referencing memory locations for client code 
leaves the memory layout \emph{ideally} in a form of DAG (directed acyclic graph) with a unique root, which resembles the first stack pointer referencing the allocated memory. DAG abstraction instead of recursive decent tree allows non-duplication on the reference counts for the identical memory objects.
% To ease picturing the memory layout after a set of requests from the client code (for allocation, referencing etc.), we can think of the unique roots of DAGs as 
% \emph{super} block addresses -- this time multiple ones, not only two -- in Figure \ref{fig:cowindex}, 
% from which a path is created for a stack pointer reference to the allocated memory. 
%\todo[inline,color=cyan]{
%Why is this a DAG? Normally programs can easily create cycles, so the DAG bit requires a little more explanation.
%Do we mean that it is ``often'' the case that it's a DAG?}

@@

\paragraph{Shallow Modalities in the Iris Separation Logic}
\label{sec:modal_in_iris}
Classically modal operators are given direct semantics in terms of an underlying model, 
which for \textsf{Iris}~\cite{jung2018iris} would be a collection of \textsc{CMRA}s ---
an algebraic cousin of a BI-algebra~\cite{ohearn1999bunched}.
This is the case for \textsf{Iris}'s one built-in modality for step-indexing~\cite{nakano-00}.
However in general, many modalities can be given proof-theoretic semantics directly within a logic. 
This is the case for \textsf{Iris}'s weakest precondition modality (which is defined within the \textsf{Iris} 
logic itself), as well as for a wide range of other modalities in the 
literature~\cite{restall2002introduction}. These modalities are typically multi-modal, 
with a form of roughly $M~\overline{ctxt}~\phi\triangleq\ldots\rightarrow \phi~\overline{ctxt}$. 
In a substructural setting~\cite{restall1993modalities,krebbers2017essence} the implication $\rightarrow$ 
is typically replaced with substructural implication $\sepimp$.

This definition also reveals a subtlety about the denotation of assertions $\phi$ --- they are effectively \emph{predicates} of some kind, functions from some possible world into some logical algebra such as a BI-algebra~\cite{ohearn1999bunched}. 
Taking this interpretation literally gives rise to a convenient way to embed new modalities in \textsf{Iris} 
in a way well-suited to many use cases. 
If $B$ is a type with a BI-algebra structure, then any space of predicate functions (e.g., $\mathsf{val}\rightarrow B$ as a simplification) also carries a BI-algebra structure, lifted from $B$, often called the 
\emph{pointwise lifting} of the algebra. A number of 
projects for for systems software verification now take advantage of this,
explicitly or implicitly lifting the BI-algebra structure from \textsf{Iris} itself to \textsf{Iris} predicates 
used as assertions within modalities. 
Connecting this to \textsf{Iris}'s typeclasses for interactive proofs~\cite{krebbers2018mosel,krebbers-17} then permits working within the embedded logic as in \textsf{Iris}.

@@

\subsection{Decomposing a System into its Constituents \emph{Contingently}}
\label{sec:definitions}
\begin{table*}[t]\small
\centering
\caption{Modal Decomposition of Program-Logics.}
\label{table:decomposition}
 %$\{\Diamond (P) \}$ write  $\{ P\}_{n}$
%$\{\overset{ICut^{n}}{\hookrightarrow} (P) \}$ return $\{ P\}_{n} $
\begin{tabular}{@{}lcccp{4cm}@{}}
  \toprule
{ Modality}& { Context} & { Elements}  &  { Nominalization} & { Context Steps} \\ \midrule
 Post-Crash~\cite{tejthesis,perennialgit,tejperennial19} &  $\Diamond \; P $  &  $  \ell \mapsto_{n}^{\overline{\gamma}} v $ &  Strong &   Crash Recovery    \\ 
%GC-Modality &      & \multicolumn{2}{c}{\multirow{2}{*}{Because why not}} & 8                                   & & ~\cite{}\\
 NextGen~\cite{larsnextgen25} &   $\overset{t}{\hookrightarrow} \; P$  &  Own (t(a))  &  Strong &  Determined Based on the Model$^{*}$   \\
 StackRegion$^{*}$  ~\cite{larsnextgen25} &  $\overset{ICut^{n}}{\hookrightarrow} \; P$      &  $\fbox{n} \; \ell \mapsto v$                                 &  Strong &  Alloc and Return to/from stack\\
 Memory-Fence~\cite{fsl,fsl++,derekrustbelt20} &  $\triangle_{\pi}$ and $\triangledown_{\pi}$      &  $\ell \mapsto v$                                 &  Weak &  Fence Acquire and Release  \\
 Address-Space~\cite{kuru2025modalabstractionsvirtualizingmemory} &  [r]P     &   $\ell \mapsto v$ &  Weak                         &  Address-Space Switch  \\ 
 Ref-Count~\cite{amalreal2024}& @$_{\ell}$ P    &  $\ell_1 \mapsto v$ &  Weak  &  Allocating, Dropping and Sharing a Reference \\ \bottomrule
\end{tabular}

{*The StackRegion Modality is an instance of NextGen (called the Independence Modality in \cite{larsnextgen25})}.
%and the $\Diamond$ Figure for Post-Crash Modality uses notationis taken from NextGen Modality paper ~\cite{larsnextgen25}.
\end{table*}

@@

\begin{definition}[Resource Elements]
%Inherited from separation logic, we identify the resource element as the indivisible logical element on which contingency defined by the modality holds. 
The simplest atomic facts we want to work with in a particular setting, specific to that setting.
\end{definition}
By definition, the resource elements are specific to some limited domain or setting.
For example, knowing that a certain address points to a 32-bit signed integer representing 3 is knowledge restricted to a certain address space.
In general,  we call these domains that any resource element is tied to \emph{resource contexts}:
\begin{definition}[Resource Context]
A resource context is an abstraction, context, or container of resource elements of the same type, e.g., an address space of a process.
\end{definition}
We discuss a range of examples for each of these in turn.

Table \ref{table:decomposition} gives additional examples of systems and their corresponding resource elements and contexts where these elements reside, though none of the work in that table analyzes itself according to the structure we are giving.
